abxD01 Paper

### 2015-04-04 PAPER.Rmd
Add flag to end of delay paragraph
Remove package loading code
Remove excess code

load_package <- function(package){
    if(!(package %in% rownames(installed.packages()))){
        install.packages(package)
    }
    library(package, quietly=TRUE)
}

load_package("randomForest")
load_package("png")

@@

***Allowing recovery of specific bacterial populations restores colonization resistance.***
In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hrs before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored [refs]. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover an additional five days. Among the metronidazole-treated mice, those that recovered longer had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)` reduction in colonization (P<0.001; Figure 4). In addition, of the `r nrow(metro_delay)` mice that experienced the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 4). Most notable among these was an OTU associated with the the family *Porphyromonadaceae* (OTU 3) and the genus *Escherichia* (OTU 4). The relative abundance of the former increased with the delay and the relative abundance of the latter decreased. Based on the correlation analysis, the trajectories of the relative abundances for these OTUs matched the observed colonization levels. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in colonization (P=`r round(amp_test$p.value, 2)`). Again, we observed a similar increase and decrease in relative abundance for the *Porphyromonadaceae* (OTU 3) and *Escherichia* (OTU 4) OTUs.

@@

--------

![**Figure 5AB.**](results/figures/Figure5/figure5ab.png)

![**Figure 5C.**](results/figures/Figure5/figure5c.png)

--------

> ### Evaluating cutoffs for OTU inclusion in Random Forest model

```{r, eval=TRUE}

dataFile <- "~/Documents/Github/abxD01/results/figures/Figure5/abxD01.final.an.unique_list.0.03.subsample.shared.toptitdel.noNewUntr.logtrans.filter16mintot.grouped.highvancD0.csv"

data<-read.csv(file=dataFile, header=T)

means <- as.data.frame(matrix(nrow=nlevels(data$expgroup), ncol=0))
row.names(means) <- levels(data$expgroup)
for(i in 3:length(data)){

  df <- aggregate( data[,i] ~ data$expgroup, data=data, mean )
  names(df)[2] <- names(data)[i]

  means <- as.data.frame(cbind( means, df[,2]))
  names(means)[dim(means)[2]] <- names(data)[i]
}

means <- rbind(means, max=-1)
for(j in 1:length(means)){
  means["max", j] <- max(means[,j])
}

means <- rbind(means, relabund=-1)
for(k in 1:length(means)){
  means["relabund", k] <- (means[17,k])/1625
}

means <- rbind(means, percent=-1)
for(k in 1:length(means)){
  means["percent", k] <- (means[18,k])*100
}

meanOTU <- means[,-1] #remove the nextDayCFU column

set <- NULL
set <- seq(0,3, by=0.01)
numOTUxThreshold <- NULL
numOTUxThreshold <- as.data.frame(cbind(numOTUxThreshold, set))
names(numOTUxThreshold)[1] <- "threshold"
numOTUxThreshold <- cbind(numOTUxThreshold, numOTU=NA)
for( i in 1:(length(set)) ) {

  test <- meanOTU[ , meanOTU[19,] >= set[i] ]
  #test2 <- sum( meanOTU[19,] >= set[i] )

  numOTUxThreshold[i, 2] <- length(test)

plot(numOTUxThreshold$numOTU ~ numOTUxThreshold$threshold, xlab="Avg % RelAbund Threshold", ylab="# OTUs included")
abline(v=seq(0,3, by=.1), col="light gray")
abline(h=seq(1,300, by=10), col="light gray")
```

> This feeds into the next analysis section.
>
> ### Random forest results

```{r, eval=TRUE}

otusAtPercAbund <- function(percThresh, meanOTU){
  perc <- meanOTU[ , meanOTU[19,] >= percThresh ]
  otus <- names(perc)[-(length(perc))]
  return(otus)
}

perc<-1
otus <- otusAtPercAbund(perc, meanOTU)

all <- read.csv(file=dataFile, header=T)
cols<- NULL
for(i in 1:(length(otus))){
  cols <-c(cols, which(names(all)==otus[i]))
}
cols <- c(which(names(all)=="Group"), which(names(all)=="nextDayCFU"), cols)

toptitdel <- all[,cols]
library(randomForest)

toptitdel.rf <- randomForest(nextDayCFU ~ .,
                           data = toptitdel[,-1],  outscale=TRUE,
                           importance=TRUE, proximity=TRUE,
                           keep.forest=TRUE, ntree=5000
)
print(toptitdel.rf)
plot(toptitdel.rf)
varImpPlot(toptitdel.rf, type=1)
imp<-importance(toptitdel.rf)
imp1<-imp[order(imp[,1], decreasing = TRUE),]
topimp <- imp1[1:15,]
write.table(topimp, file="~/Desktop/mothur/abxD01/rf/rf.toptitdel.1p.importance.txt", sep="\t", row.names=TRUE)

> ### Figure 5A: Evaluating and plotting the fit of the model

```{r, eval=FALSE}
predictions <- predict(toptitdel.rf, newdata=toptitdel[,-c(1,2)])
actual <- toptitdel$nextDayCFU
expgroup<- as.character(all[,2])
results<-as.data.frame(cbind(actual, predictions))
colnames(results) <- c( "actual", "predict")

ybar <- apply(results, 2, mean)
num<-sum((results$actual-ybar["actual"])*(results$predict-ybar["predict"]))
denA <- sum((results$actual-ybar["actual"])^2)
denB <- sum((results$predict-ybar["predict"])^2)
rsq <- (num^2)/(denA*denB) #calculated from the square of the sample correlation coefficient, little r squared as opposed to big R squared

fit.results <- cbind(expgroup, results)

#Plot observed vs predicted
#information for plotting
color<-rainbow(7)
colors <- c(control = "black",
          cipro = color[1],
          clinda = "#FFCC00",
          vanc.625 = "#00CC00",
          vanc.3 = "#00CC00",
          vanc.1 = "#00CC00",
          strep5 = "#33FFFF",
          strep.5 = "#33FFFF",
          strep.1 = "#33FFFF",
          cef.5 = color[5],
          cef.3 = color[5],
          cef.1 = color[5],
          amp.5 = "#6600FF",
          amp.5d = "#6600FF",
          metro1 = color[7],
          metro1d = color[7])

# topdose is all circles, middle dose is square,
# low dose is upward pointing triangle, delayed is starburst
pch <- c(control = 1,
         cipro = 16,
         clinda = 16,
         vanc.625 = 16,
         vanc.3 = 15,
         vanc.1 = 17,
         strep5 = 16,
         strep.5 = 15,
         strep.1 = 17,
         cef.5 = 16,
         cef.3 = 15,
         cef.1 = 17,
         amp.5 = 16,
         amp.5d = 8,
         metro1 = 16,
         metro1d = 8)

#these are needed for the legend
legend.labels <- c(control = "Control",
            cipro = "Ciprofloxacin",
            clinda = "Clindamycin",
            vanc.625 = "Vancomycin 0.625 mg/ml",
            vanc.3 = "Vancomycin 0.3 mg/ml",
            vanc.1 = "Vancomycin 0.1 mg/ml",
            strep5 = "Streptomycin 5 mg/ml",
            strep.5 = "Streptomycin 0.5 mg/ml",
            strep.1 = "Streptomycin 0.1 mg/ml",
            cef.5 = "Cefoperazone 0.5 mg/ml",
            cef.3 = "Cefoperazone 0.3 mg/ml",
            cef.1 = "Cefoperazone 0.1 mg/ml",
            amp.5 = "Ampicillin",
            amp.5d = "Ampicillin +5 days",
            metro1 = "Metronidazole",
            metro1d = "Metronidazole +5 days")

design<-as.matrix(1:2)
design<-cbind(design, c(3,3))
layout(design, widths=c(1,1), heights=c(.35, 1))

plot.new()
par(mar=c(0.5, 0.5, 0.5, 0.5))
legend("bottomright", legend=legend.labels, pch=pch, col=colors, pt.cex=1.5,  cex=1, bty="n", ncol = 2, text.width=0.4)

par(mar=c(5, 5, 0.5, 0.5) +0.1, mgp=c(3, 1, 0), las=1)
plot(fit.results[fit.results[,1]=="control",2],
     fit.results[fit.results[,1]=="control",3],
     main="",
     ylab=expression(paste("Predicted Log ", italic("C. difficile"), " CFU/g Feces")),
     xlab=expression(paste("Actual Log ", italic("C. difficile"), " CFU/g Feces")),
     ylim=c(0,9), xlim=c(0,9), xaxt='n', yaxt='n', cex=1.5)
abline(a=0, b=1, lty="dashed", lwd=2, col="black")
axis(1, at = c(0:9), labels=c(0:9))
axis(2, at = c(0:9), labels=c(0:9))

for(i in 2:(length(legend.labels))){
  points(fit.results[fit.results[,1]==names(legend.labels)[i],2],
         fit.results[fit.results[,1]==names(legend.labels)[i],3],
         col=colors[i], pch=pch[i], cex=1.5)
}

> ### Figure 5B: Importance Plot

```{r, eval=FALSE}
topimp<-topimp[order(topimp[,1], decreasing=FALSE),]
ids <- read.csv(file = "~/Documents/Github/abxD01/Figure5/rf.toptitdel.1p.importance.ids.csv", header = TRUE)
#par(mfrow=c(1, 1)) #+1 to give extra labeling space
par(mar=c(5, 15, 0.5, 1) +0.1, mgp=c(3, 1, 0), las=0) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)
plot(topimp[,1], 1:15, xlab="% Increase in Mean Squared Error", yaxt="n", ylab="", pch=16, cex=1.5, xlim=c(20, 80))
axis(2, at = c(1:15), labels = rep("", 15), las=1, cex.axis=.9 )
for(z in 1:(dim(ids)[1])){
  name<-as.character(ids$name[z])
  otuname<- as.character(ids$otuname[z])
  mtext(text = bquote(italic(.(name))*" ("*.(otuname)*")"), at=z, side = 2, line = 1, las=2, cex=0.9)
}
abline(h=c(1:15), lty="dashed", col='black')

> ### Testing a random forest model with only OTUs 3 and 39

```{r, eval=TRUE}
test3.39 <- toptitdel[,c("Group","nextDayCFU","Otu00003", "Otu00039")]

rf.test3.39 <- randomForest(nextDayCFU ~ .,
                           data = test3.39[,-1],  outscale=TRUE,
                           importance=TRUE, proximity=TRUE,
                           keep.forest=TRUE, ntree=5000
)
print(rf.test3.39)
plot(rf.test3.39)

predictions <- predict(rf.test3.39, newdata=test3.39[,-c(1,2)])
actual <- test3.39$nextDayCFU
test.results<-as.data.frame(cbind(actual, predictions))
colnames(test.results) <- c( "actual", "predict")

ybar <- apply(test.results, 2, mean)
num<-sum((test.results$actual-ybar["actual"])*(test.results$predict-ybar["predict"]))
denA <- sum((test.results$actual-ybar["actual"])^2)
denB <- sum((test.results$predict-ybar["predict"])^2)
test.rsq <- (num^2)/(denA*denB)

> ### Correlation over all three datasets with *C. difficile* levels

```{r, eval=FALSE}

#First the correlation calculations that were put on the graph:
c<-1
otu <- c()
cor.spear <- c()
pval.spear <- c()
cor.ken <- c()
pval.ken <- c()
for(i in 3:length(toptitdel)){
  otu[c] <- colnames(toptitdel[i])
  cor.spear[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="spearman")$estimate
  pval.spear[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="spearman")$p.value
  cor.ken[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="kendall")$estimate #good to see because kendall handles ties
  pval.ken[c] <- cor.test(toptitdel[,2],toptitdel[,i], method="kendall")$p.value #but only works if this is tao-b and not tao-a which im not sure about
  c <- c+1
}
pval.spear<-p.adjust(pval.spear, method='BH') #adjust for multiple comparisons
pval.ken<-p.adjust(pval.ken, method='BH') #adjust for multiple comparisons

correl = NULL
correl <- matrix(c(otu, cor.spear, pval.spear, cor.ken, pval.ken), ncol=5)
colnames(correl) <- c('otu','corSpear','pvalSpear', "corKen", "pvalKen")
correl <- correl[order(correl[,3]),]  #order by pvalue column=3
#write.table(correl, file="~/Desktop/mothur/abxD01/correlation/toptidel.1p.correl.txt", sep="\t", row.names=FALSE)
print(correl)
```

> From this file I selected the top 8 OTUs identified by random forest's increase in mean squared error measurement to graph (next section).
>
> ### Figure 5C: OTU abundance vs. *C. difficile* abundance

```{r, eval=FALSE}
topimp.otus <-  row.names(topimp)[c(15:7)]
temp.match <- match(topimp.otus, ids.master[,1])
graphID <- ids.master[temp.match,c(4,3)]

temp.match <- match(topimp.otus, correl[,"otu"])
topimp.corrs <- correl[temp.match, c("corSpear", "pvalSpear")]
row.names(topimp.corrs) <- topimp.otus

graphOTUxCD(otus = topimp.otus, ids = graphID, corrs = topimp.corrs)
```

>>>>

@@

***Allowing recovery of specific bacterial populations restores colonization resistance.***
In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hrs before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored [refs]. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover an additional five days. Among the metronidazole-treated mice, those that recovered longer had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)` reduction in colonization (P<0.001; Figure 4). In addition, of the `r nrow(metro_delay)` mice that experienced the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 4). Most notable among these was an OTU associated with the the family *Porphyromonadaceae* (OTU 3) and the genus *Escherichia* (OTU 4). The relative abundance of the former increased with the delay and the relative abundance of the latter decreased. Based on the correlation analysis, the trajectories of the relative abundances for these OTUs matched the observed colonization levels. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in colonization (P=`r round(amp_test$p.value, 2)`). Again, we observed a similar increase and decrease in relative abundance for the *Porphyromonadaceae* (OTU 3) and *Escherichia* (OTU 4) OTUs. <Need a conclusion sentence here>

@@
